\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-29 paper.qmd
Updated Figure 8

nldr_two_curvy <- umap_two_curvy_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_two_curvy_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5, size = 0.5, color = "#000000") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)

@@

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(color = "black", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             color = "#33a02c",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos) 
```

@@

trimesh_removed_two_curvy <- vis_rmlg_mesh(
  distance_edges = distance_two_curvy, 
  benchmark_value = benchmark_two_curvy, 
  tr_coord_df = tr_from_to_df_two_curvy, 
  distance_col = "distance") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("d", sc_ltr_pos)

@@

@@

two_curvy_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#000000") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.126, 0.125)) + 
  ylim(c(-0.126, 0.125)) +
  interior_annotation("a", c(0.08, 0.93), cex = 2) +
  theme(
    legend.position = "none"
  )

@@

two_curvy_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#000000") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.125, 0.075)) + 
  ylim(c(-0.125, 0.075)) +
  interior_annotation("b", c(0.08, 0.93), cex = 2) +
  theme(
    legend.position = "none"
  )
```

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two views of the true model (blue lines) in $2\\text{-}D$ projections from $7\\text{-}D$, for the two-curvy clusters data (black points). The data is spread along the non-linear shaped clusters, and does not vary much from the curves. The **langevitour** software is used to view the data with a tour, and the full video is available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5

@@

mse_two_curvy_b <- ggplot(error_two_curvy_clust_umap, 
                     aes(x = a1, 
                         y = MSE)) +
  geom_vline(xintercept = 0.09,
             linetype="solid",
           color = "#d95f02", linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1b9e77", linetype=2,
             linewidth=1) +
  geom_vline(xintercept = 0.12, linetype=2,
             color = "#1f78b4", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 3, 5, 8, 10, 11, 12, 13, 14)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "MSE") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

prop_dens_a1 <- ggplot(error_two_curvy_clust_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
             linetype="solid",
           color = "#d95f02", linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1b9e77", linetype=2,
             linewidth=1) +
  geom_vline(xintercept = 0.12, linetype=2,
             color = "#1f78b4", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 3, 5, 8, 10, 11, 12, 13, 14)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("proportion of densities")) +
  ggtitle("(d)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

@@

a1_m_two_curvy <- ggplot(error_two_curvy_clust_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
             linetype="solid",
           color = "#d95f02", linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1b9e77", linetype=2,
             linewidth=1) +
  geom_vline(xintercept = 0.12, linetype=2,
             color = "#1f78b4", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 3, 5, 8, 10, 11, 12, 13, 14)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("proportion of non-empty bins ", bgroup("(", frac(m, b), ")")))) +
  ggtitle("(b)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $4\\text{-}D$ model error in \\gD{} layout. Color indicates square root of absolute error, dark blue indicating high error and light indicates low error. Most large errors are distributed near the edges of the clusters."
#| out-height: 30%
#| fig-pos: H

@@

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the two-curvy clusters data, using three different bin inputs: (a) $b = 63 \\text{ } (9, \\text{ }7)$, (b) $b = 108 \\text{ } (12, \\text{ }9)$, and (c) $b = 312 \\text{ } (14, \\text{ }10)$. Color indicates standardized counts, dark indicating high count and light indicates low count. Using the MSE of the model fit in $7\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 9
#| fig-height: 2

>>>>

nldr_two_curvy <- umap_two_curvy_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#636363", size = 0.5) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_two_curvy_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5, size = 0.5, color = "#636363") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)

@@

hex_grid_nonempty_two_curvy <- ggplot(
  data = hex_grid_nonempty, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = df_bin_centroids_two_curvy, 
             aes(x = c_x, y = c_y), 
             color = "#000000",
             size = 1) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c", sc_ltr_pos) 
```

@@

# trimesh_removed_two_curvy <- vis_rmlg_mesh(
#   distance_edges = distance_two_curvy, 
#   benchmark_value = benchmark_two_curvy, 
#   tr_coord_df = tr_from_to_df_two_curvy, 
#   distance_col = "distance") +
#   xlim(sc_xlims) + ylim(sc_ylims) +
#   interior_annotation("d", sc_ltr_pos)

@@

trimesh_removed_two_curvy <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  coord_equal() +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("d", sc_ltr_pos)

@@

two_curvy_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.126, 0.125)) + 
  ylim(c(-0.126, 0.125)) +
  interior_annotation("a", c(0.08, 0.93), cex = 2) +
  theme(
    legend.position = "none"
  )

@@

two_curvy_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#636363") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.125, 0.075)) + 
  ylim(c(-0.125, 0.075)) +
  interior_annotation("b", c(0.08, 0.93), cex = 2) +
  theme(
    legend.position = "none"
  )
```

```{r}
#| label: fig-two-curvy-true-proj
#| fig-cap: "Two views of the true model (blue lines) in $2\\text{-}D$ projections from $7\\text{-}D$, for the two-curvy clusters data (grey points). The data is spread along the non-linear shaped clusters, and does not vary much from the curves. The **langevitour** software is used to view the data with a tour, and the full video is available at <>." 
#| fig-pos: H
#| fig-width: 10
#| fig-height: 5

@@

mse_two_curvy_b <- ggplot(error_two_curvy_clust_umap, 
                     aes(x = a1, 
                         y = MSE)) +
  geom_vline(xintercept = 0.09,
             linetype="solid",
           color = "#d95f02", linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1b9e77", linetype=2,
             linewidth=1) +
  geom_vline(xintercept = 0.12, linetype=2,
             color = "#1f78b4", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 4, 8, 10, 11, 12, 13, 14)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "MSE") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

prop_dens_a1 <- ggplot(error_two_curvy_clust_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
             linetype="solid",
           color = "#d95f02", linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1b9e77", linetype=2,
             linewidth=1) +
  geom_vline(xintercept = 0.12, linetype=2,
             color = "#1f78b4", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 4, 8, 10, 11, 12, 13, 14)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("proportion of densities")) +
  ggtitle("(d)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

@@

a1_m_two_curvy <- ggplot(error_two_curvy_clust_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
             linetype="solid",
           color = "#d95f02", linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1b9e77", linetype=2,
             linewidth=1) +
  geom_vline(xintercept = 0.12, linetype=2,
             color = "#1f78b4", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_clust_umap$a1, 2)))[c(1, 4, 8, 10, 11, 12, 13, 14)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("proportion of non-empty bins ", bgroup("(", frac(m, b), ")")))) +
  ggtitle("(b)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line())

@@

```{r}
#| label: fig-p-d-error-in-2d-two-curvy
#| fig-cap: "The $7\\text{-}D$ model error in \\gD{} layout. Color indicates square root of absolute error, dark blue indicating high error and light indicates low error. Most large errors are distributed near the edges of the clusters."
#| out-height: 30%
#| fig-pos: H

@@

```{r}
#| echo: false
#| label: fig-bins-two-curvy
#| fig-pos: H
#| fig-cap: "Hexbin density plots of UMAP layout of the two-curvy clusters data, using three different bin inputs: (a) $b = 80 \\text{ } (10, \\text{ }8)$, (b) $b = 130 \\text{ } (13, \\text{ }10)$, and (c) $b = 391 \\text{ } (23, \\text{ }17)$. Color indicates standardized counts, dark indicating high count and light indicates low count. Using the MSE of the model fit in $7\\text{-}D$ helps decide on a useful choice of number of bins."
#| fig-width: 9
#| fig-height: 2

### 2024-09-29 paper.tex
Updated Figure 8

\caption{\label{fig-two-curvy-true-proj}Two views of the true model
(blue lines) in \(2\text{-}D\) projections from \(7\text{-}D\), for the
two-curvy clusters data (black points). The data is spread along the
non-linear shaped clusters, and does not vary much from the curves. The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at \textless\textgreater.}

@@

\caption{\label{fig-p-d-error-in-2d-two-curvy}The \(4\text{-}D\) model
error in \gD{} layout. Color indicates square root of absolute error,
dark blue indicating high error and light indicates low error. Most
large errors are distributed near the edges of the clusters.}

@@

\caption{\label{fig-bins-two-curvy}Hexbin density plots of UMAP layout
of the two-curvy clusters data, using three different bin inputs: (a)
\(b = 63 \text{ } (9, \text{ }7)\), (b)
\(b = 108 \text{ } (12, \text{ }9)\), and (c)
\(b = 312 \text{ } (14, \text{ }10)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. Using
the MSE of the model fit in \(7\text{-}D\) helps decide on a useful
choice of number of bins.}

>>>>

\caption{\label{fig-two-curvy-true-proj}Two views of the true model
(blue lines) in \(2\text{-}D\) projections from \(7\text{-}D\), for the
two-curvy clusters data (grey points). The data is spread along the
non-linear shaped clusters, and does not vary much from the curves. The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at \textless\textgreater.}

@@

\caption{\label{fig-p-d-error-in-2d-two-curvy}The \(7\text{-}D\) model
error in \gD{} layout. Color indicates square root of absolute error,
dark blue indicating high error and light indicates low error. Most
large errors are distributed near the edges of the clusters.}

@@

\caption{\label{fig-bins-two-curvy}Hexbin density plots of UMAP layout
of the two-curvy clusters data, using three different bin inputs: (a)
\(b = 80 \text{ } (10, \text{ }8)\), (b)
\(b = 130 \text{ } (13, \text{ }10)\), and (c)
\(b = 391 \text{ } (23, \text{ }17)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. Using
the MSE of the model fit in \(7\text{-}D\) helps decide on a useful
choice of number of bins.}
